---
title: "Covid"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(roll)
library(magrittr)

# import data frames COVID: cases, tests and population size of municipalities
dc <- read_delim(here("exercises", "ws02", "covid","data", "Municipality_cases_time_series.csv"), ";", escape_double = FALSE, trim_ws = TRUE)
dt <- read_delim(here("exercises", "ws02", "covid","data", "Municipality_tested_persons_time_series.csv"), ";", escape_double = FALSE, trim_ws = TRUE)
dsize <- read_delim(here("exercises", "ws02", "covid","data", "Municipality_test_pos.csv"), ";", escape_double = FALSE, trim_ws = TRUE)
```
## Covid data analysis DK overview
```{r dataFramesGlimpse , include=TRUE}
dc
dt
dsize
```

```{r dc-diagnosedCases, include=TRUE}
# make data tidy
dc %<>% 
  #use pivot_longer to transpose all columns (apart from date_sample) to two columns ('kommune','testsConducted') that hold the name of the kommune and the values of the number of diagnosed cases
  pivot_longer(cols = !date_sample, names_to = "kommune", values_to = "casesDiagnosed") %>% arrange(kommune, date_sample)
dc
```

```{r testedPeople, include=TRUE}
dt %<>% 
  #use pivot_longer to transpose all columns (apart from PrDate_adjusted) to two columns ('kommune','testsConducted') that hold the name of the kommune and the values of the number of conducted tests 
  pivot_longer(cols = !PrDate_adjusted, names_to = "kommune", values_to = "testsConducted") %>%
  arrange(kommune, PrDate_adjusted) %>%
  rename(date_sample = PrDate_adjusted) %>%
  filter(!(kommune == "X101"))

```

```{r NAtests, include=TRUE}
# check there are no NA in the sets dc and dt , dc$casesDiagnosed, and dt$testsConducted
sum(is.na(dc$casesDiagnosed))
sum(is.na(dt$testsConducted))
sum(is.na(dsize$Befolkningstal))
```

```{r columnRenaming, include=TRUE}
# make data compatible rename columns for merging 
dsize <- dsize %>%
  select(contains("Kom"), Befolkningstal) %>%
  rename("kID" = `Kommune_(id)`, "kommune" = `Kommune_(navn)`, population = Befolkningstal) %>%
  mutate(population = population * 1000) %>%
  select(-kID)

```

```{r keyCheck, include=TRUE}
# check Kommune name can be used as key
dc[!(dc$kommune %in% dsize$kommune), ]$kommune
dsize[!(dsize$kommune %in% dt$kommune), ]$kommune
dt[!(dt$kommune %in% dsize$kommune), ]$kommune
```

```{r Cph-kbh-CleanUp, include=TRUE}
# ooops!
# for some reason one data frame uses Københaven the other Copenhagen), - use str_replace to change "København", "Copenhagen"
dsize$kommune <- str_replace(dsize$kommune, "København", "Copenhagen")

# check again keys are clean OK re-run rows 35-37

# merge data together
dc <- merge(dc, dsize)
dt <- merge(dt, dsize)

```

```{r DeriveVariablesCases, include=TRUE}

# create new variables within mutate 
# (e.g. rolling aggregates with roll_sum - cases of the last seven days) - 
# numbers per 1000 people 
# use the lag function to compute the differences between a days (7day sum) value and the one from 7 days (3, and 1) ago
#
dc %<>% 
  group_by(kommune) %>%
  arrange(kommune,date_sample) %>%
  mutate(dcr100k = casesDiagnosed / (population / 100000), 
         dcr7d = roll_sum(casesDiagnosed, width = 7, min_obs = 1) / (population / 100000),
         dcr7dPer100k= dcr7d /(population / 100000),
         dcr7dPer100kCh1 = lag(dcr7dPer100k,1)-dcr7dPer100k,
         dcr7dPer100kCh3 = lag(dcr7dPer100k,3)-dcr7dPer100k,
         dcr7dPer100kCh7 = lag(dcr7dPer100k,7)-dcr7dPer100k
         )

dc
```

```{r DeriveVariablesTests}
dt %<>% 
  group_by(kommune) %>% 
  arrange(kommune,date_sample) %>%
  mutate(tr100k = testsConducted / (population / 100000), 
         tr7d = roll_sum(testsConducted, width = 7, min_obs = 1), 
         tr7dP100k = tr7d / (population / 100000),
         tr7dP100kCh1 = lag(tr7dP100k,1)-tr7dP100k,
         tr7dP100kCh3 = lag(tr7dP100k,3)-tr7dP100k,
         tr7dP100kCh7 = lag(tr7dP100k,7)-tr7dP100k
         )
dt
```

```{r Merge}
# merge test with case data (on shared keys - date_sample+kommune)
df <- merge(dc, dt)
df

#create test positive variables 
df %<>% 
  mutate(
    PosTestRateOnDay = dcr100k / tr100k,
    PosTestRate7dp100k = dcr7dPer100k / tr7dP100k, 
  )
    

# check there were no diagnosis without tests
sum(df$casesDiagnosed > df$testsConducted)
```

## Covid data analysis DK overview
df%>% select(PosTestRate7dp100k, dcr7dPer100k,kommune,date_sample) %>% filter(kommune=="Aalborg") %>% arrange(date_sample) %>% ggplot(aes(PosTestRate7dp100k,dcr7dPer100k))+geom_line() +geom_segment(aes(
                    xend=c(tail(PosTestRate7dp100k, n=-1), NA), 
                    yend=c(tail(dcr7dPer100k, n=-1), NA)
                  )
      ) 

df%>% select(PosTestRate7dp100k, dcr7dPer100k,kommune,date_sample) %>% filter(kommune=="Aalborg") %>% arrange(date_sample) %>% ggplot(aes(PosTestRate7dp100k,dcr7dPer100k,label=date_sample))+geom_point()+geom_line() +geom_segment(aes(
                    xend=c(tail(PosTestRate7dp100k, n=-1), NA), 
                    yend=c(tail(dcr7dPer100k, n=-1), NA)
                  ), arrow=arrow(length=unit(0.3,"cm"))
      ) 



df%>% filter(kommune=="Aalborg") %>% ggplot(aes(date_sample,dcr7dPer100k))+geom_line()

df%>% filter(kommune=="Aalborg") %>% ggplot(aes(date_sample,PosTestRate7dp100k))+geom_line()

